---
title: "Getting started with RAILS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with RAILS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(RAILS)
```

## The problem

A volunteer biobank recruits whoever signs up. The people who sign up are not a
random draw from the population: they skew older, wealthier, better educated,
and — the part that matters — those characteristics are associated with the
health outcomes the biobank exists to study. Any prevalence estimated directly
from the cohort is biased for the population.

A probability survey, by contrast, is designed. Its design weights are known,
and its margins represent the population. It is usually far smaller than the
biobank and does not measure what the biobank measures.

RAILS uses the second to fix the first.

## A population where we know the answer

Real biobank microdata cannot be redistributed, so this vignette uses
`rails_simulate()`, which generates a population, a probability sample, and a
volunteer sample whose inclusion depends on the same covariates that drive the
outcome. Because we simulated the population, we know the truth.

```{r simulate}
pop <- rails_simulate(20000, seed = 42)

aou <- pop[pop$aou == 1, ]        # the volunteer sample
ref <- pop[pop$s == 1, ]          # the probability sample
ref$dweight <- 1 / ref$ps         # its design weights

c(population = nrow(pop), volunteer = nrow(aou), reference = nrow(ref))
```

The bias is visible before any modelling. The volunteer sample over-represents
the higher income groups:

```{r bias}
round(rbind(
  population = prop.table(table(pop$income)),
  volunteer  = prop.table(table(aou$income))
), 3)
```

and its raw outcome prevalence is off:

```{r bias-y}
c(truth = mean(pop$y), unweighted = mean(aou$y))
```

## Fitting

`rails()` takes the two samples, the covariates they share, and the reference
sample's design weights. Two further arguments say which models the search runs
between: `start` is the model taken as given, `scope` the largest model
selection may reach.

Here the volunteer-selection process contains two-way structure, so we start
from main effects and let the forward test look for the two-way interactions:

```{r fit}
fit <- rails(
  np   = aou,
  ref  = ref,
  vars = c("agegroup", "sex", "income"),
  weights_ref = "dweight",
  start = 1,          # main effects, taken as given
  scope = 2,          # two-way interactions are the candidates
  verbose = FALSE
)

fit
```

Three things happened. A propensity model was fitted on the starting model.
Every term in `scope` but not in `start` was offered to a forward
likelihood-ratio test, and those passing at `alpha = 0.05` were kept. Then the
selected terms were added back one at a time, raking to the reference margins at
each step and stopping at the first step that failed to converge.

The defaults are `start = 2, scope = 3`, which is the published estimator:
main effects and two-way interactions taken as given, three-way interactions
offered for selection.

`summary()` shows the selection path and how well behaved the weights are:

```{r summary}
summary(fit)
```

The design effect is the price of weighting: it is the factor by which unequal
weights inflate the variance of a mean, so an effective sample size well below
the nominal one is normal and expected. A design effect that is very large
signals a handful of records carrying the estimate, which is worth looking at:

```{r plot, fig.width = 6, fig.height = 4}
plot(fit)
```

## Using the weights

```{r weights}
w <- weights(fit)

c(truth      = mean(pop$y),
  unweighted = mean(aou$y),
  rails      = sum(w * aou$y) / sum(w))
```

The weights sum to the population total implied by the reference sample, and
reproduce its margins by construction:

```{r margins}
round(rbind(
  reference = tapply(ref$dweight, ref$income, sum),
  rails     = tapply(w, aou$income, sum)
))
```

## Standard errors

`rails_var()` gives the standard error and interval for the weighted mean. By
default it reports the *simplified* variance, which treats the weights as fixed:

```{r var}
round(rails_var(fit, aou$y, truth = mean(pop$y)), 5)
```

But the weights were estimated, not given. The *stacked* variance accounts for
that, through a sandwich over three estimating equations: the propensity score,
the raking constraints, and the mean itself. Ask for both to see the gap:

```{r var-both}
v <- rails_var(fit, aou$y, type = "both", truth = mean(pop$y))
round(v[c("mu_hat", "se_naive", "se_full", "ratio_full_naive")], 5)
```

`ratio_full_naive` is the ratio of the two. Note that it is below 1 here: the
stacked figure is the *smaller* one. Ignoring the propensity stage tends to
understate, but raking to known population margins removes variance the way a
regression estimator does, and in this example the second effect wins. The
stacked form is the one to report because it accounts for how the weights were
built -- not because it is the more conservative number.

The corresponding interval:

```{r ci}
round(v[c("ci_full_lower", "ci_full_upper")], 4)
```

The simplified form is the cheaper of the two -- it needs no model matrix at all
-- which is why it is the default. Both run on the aggregated cells rather than
on individual records, so neither needs `keep_data = TRUE`, and both work on a
fit built with `aggregated = TRUE` when you pass the outcome per cell as
`sum` and `sumsq`.

## Downstream analysis

For anything beyond a mean, hand the weights to **survey**:

```{r design}
des <- rails_design(fit, aou)
survey::svymean(~y, des)
```

Bear in mind that standard errors from this object are the naive ones, for the
same reason as above. Use it for point estimates and model fitting; use
`rails_var()` for inference on a mean.

## Where to go next

- `vignette("subgroup-rails")` — running the estimator within strata, when
  subgroup estimates are the goal.
- `?rails` — the full argument list, including `start` / `scope` for which models
  the search runs between, and `aggregated` for pre-built cell tables.
- `?rails_cells` and `?rails_totals` — building the aggregated inputs yourself,
  which is what you want when the microdata will not fit in one data frame.
- `?pums_reference` — building the reference sample from ACS PUMS instead of
  from a simulation: `pums_fetch()` downloads it, `pums_recode()` bands the raw
  ACS codes into these same analysis variables, and `pums_reference()` returns
  the cleaned records, the cell table and the margins together.
