---
title: "Subgroup RAILS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Subgroup RAILS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(RAILS)
```

## Why not just subset

Suppose the question is not national prevalence but prevalence *by region*, or
*by sex*. The obvious move is to fit `rails()` once on the whole sample and then
sum the weights within each subgroup.

That is not the same estimator, and it is weaker in two ways. A single global fit
selects one set of interactions for the whole population, so a term that matters
only in one stratum has to be strong enough to survive a test run on everyone.
And a global fit rakes to national margins, which constrains the weights to
reproduce national totals — not each subgroup's own totals.

`rails_subgroup()` instead runs the whole procedure independently within each
level: its own propensity model, its own selection path, its own stepwise walk,
and its own population margins built from that stratum's reference records.

## Fitting

```{r simulate}
pop <- rails_simulate(20000, seed = 7)

aou <- pop[pop$aou == 1, ]
ref <- pop[pop$s == 1, ]
ref$dweight <- 1 / ref$ps
```

`by` names the stratifier. It must be a column of both samples, and it must not
also appear in `vars` — within a stratum it is constant, so it cannot be a
covariate there.

### Choosing a stratifier

Which variable you stratify by is worth a moment's thought, because it decides
what is left for the selection step to find.

The reason is that conditioning removes structure. Whatever you hold fixed can no
longer interact with anything: an interaction between the stratifier and another
variable does not disappear, it simply reappears as a main effect inside each
stratum, where there is nothing left to detect.

That matters here because this synthetic population is built with two
interactions driving who volunteers — age with sex, and sex with income — and
nothing else. Hold sex fixed and both of them flatten: age×sex becomes a plain
age effect, sex×income becomes a plain income effect, and the only candidate the
forward test could still consider inside a stratum, age×income, was never in the
population to begin with. Every stratum would correctly select nothing, and the
vignette would demonstrate nothing.

Hold *age* fixed instead and sex and income both still vary, with the sex×income
term intact in each stratum. So that is what the fit below does: stratify by
`agegroup`, and let the selection step look for interactions among `sex` and
`income`.

None of this is a limitation of the method — it is a property of how this
particular example was simulated. But the same question is worth asking of real
data: if the structure you care about runs through the variable you are
stratifying by, you will not find it inside the strata.

```{r fit}
fit <- rails_subgroup(
  np   = aou,
  ref  = ref,
  vars = c("sex", "income"),
  by   = "agegroup",
  weights_ref = "dweight",
  start = 1,          # as in the getting-started vignette
  scope = 2,
  fallback = TRUE,    # explained below
  verbose = FALSE
)

fit
```

Note in the printout that the strata need not select the same terms. That is the
point of the method: the covariate structure driving volunteer participation can
genuinely differ between subpopulations, and forcing one model on all of them
gives up that information. Here two strata pick up `sex:income` and the third
does not — the term is real, but not every stratum has the sample size to detect
it.

```{r paths}
lapply(fit$fits, function(f) f$terms_selected)
```

### What happens when a stratum selects nothing

That third stratum is why `fallback = TRUE` appears in the call above, and it is
worth understanding rather than copying.

RAILS builds its weights in two stages: selection proposes terms, then those
terms are raked in one at a time. If selection proposes nothing — or if the very
first raking step fails to converge — there is no higher-order model to hand
back, and something has to be returned anyway.

By default RAILS returns `NA`, which is what the published code does. That is a
deliberate choice: it makes a non-result impossible to miss, so nobody reports
starting-model weights as though a selection step had confirmed them.

`fallback = TRUE` returns the raked starting model in that case instead. The
weights are perfectly usable — they are simply the starting model's weights,
with nothing higher-order added. The two settings on the same data:

```{r fallback}
strict  <- suppressWarnings(rails_subgroup(
  aou, ref, c("sex", "income"), by = "agegroup", weights_ref = "dweight",
  start = 1, scope = 2, verbose = FALSE))            # fallback = FALSE

c(strict  = sum(is.na(weights(strict))),
  relaxed = sum(is.na(weights(fit))))
```

Either way `converged` records which happened, so the two are always
distinguishable after the fact:

```{r converged}
vapply(fit$fits, function(f) f$converged, logical(1))
```

In a real analysis, `FALSE` is the safer default and `TRUE` is the pragmatic one
for a pipeline that has to produce a weight for every record. This vignette uses
`TRUE` so that all three strata have weights to show.

## Using the weights

The returned object carries one pooled weight vector aligned to the rows of the
non-probability sample, so it is used exactly like a global fit's:

```{r weights}
w <- weights(fit)

## Each stratum is scaled to its own reference total.
round(rbind(
  reference = tapply(ref$dweight, ref$agegroup, sum),
  rails     = tapply(w, aou$agegroup, sum)
))
```

Subgroup estimates, against the truth we happen to know:

```{r estimates}
est <- tapply(seq_along(w), aou$agegroup,
              function(i) sum(w[i] * aou$y[i]) / sum(w[i]))

round(rbind(
  truth      = tapply(pop$y, pop$agegroup, mean),
  unweighted = tapply(aou$y, aou$agegroup, mean),
  rails      = est
), 4)
```

The individual `rails_fit` objects are available if a stratum needs inspecting on
its own:

```{r one-stratum}
summary(fit$fits[[1]])
```

## Standard errors

`rails_var()` takes a subgroup fit and returns one row per stratum:

```{r var}
rails_var(fit, aou$y, truth = tapply(pop$y, pop$agegroup, mean))
```

`truth` may be a single value or, as here, one per stratum named by level.

The default is the **simplified** variance, and for subgroup RAILS that default
is the one the paper reports. Residual bias persists at the subgroup level, so a
tighter interval would overstate what the estimator actually delivers.

The **stacked** form is available and does apply within a stratum — each stratum
has its own propensity model, its own selected terms and its own margins, so the
estimating equations really are separate. What the per-stratum sandwich does not
carry is the covariance *between* strata, and it says so:

```{r var-stacked}
rails_var(fit, aou$y, type = "stacked")[, c("agegroup", "mu_hat", "se_full")]
```

So those intervals are usable one stratum at a time. Anything that pools strata,
or takes a difference between two of them, needs the between-stratum covariance
that is not computed here — bootstrap the whole procedure if you need that.

## Empty strata

A level present in the volunteer sample but absent from the reference sample
cannot be fitted: there are no margins to rake to. Those levels are skipped with
a warning and their records receive `NA` weights, rather than being silently
dropped or pooled into a neighbouring stratum. Check for them before
aggregating:

```{r na-check}
sum(is.na(w))
```
